---
title: "Exploring Data"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Loading Libraries
```{r, message=FALSE}
library(tidyverse)
```

I'm just doing a super quick look at the data right now. Going to load them in, do a simple regression, and see which ones might be the best to move forward with.


# Loading default data
```{r, message=FALSE}
defaults <- read_csv("data/default-credit-cards.csv", skip = 1)
```

# Cleaning
```{r}
# Making lower case columns
colnames(defaults) <- defaults %>% 
  colnames() %>% 
  tolower()

# Renaming default column
defaults <- defaults %>% 
  rename(default = `default payment next month`) %>% 
  select(-id)
```

# Quick initial regression
```{r}
reg <- glm(default ~ ., family = "binomial", data = defaults)

summary(reg)
```

# Loading news popularity
```{r}
news <- read_csv("data/online-news-popularity.csv")
```

# Cleaning
```{r}
# Getting rid of non-predictive variables
news <- news %>% 
  select(-url, -timedelta)
```

# Quick initial regression
```{r}
reg <- lm(shares ~ ., data = news)
summary(reg)
```

# Loading student info
```{r}
# Student math and Portuguese data
student_mat <- read_delim("data/student-mat.csv", delim = ";")
student_por <- read_delim("data/student-por.csv", delim = ";")

# Combining
student_combined <- inner_join(student_mat, student_por, by = c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
```

Okay so, for these data sets there are different numbers of observations. Math has 395, Portuguese has 649, and combined has 382. So I guess there isn't a lot of overlap between the two. Maybe it would be best to just focus on the Portuguese one if we choose this? Just because there are some more observations.

# Quick regressions
```{r}
math_reg <- lm(G3 ~ ., data = student_mat)
por_reg <- lm(G3 ~ ., data = student_por)
combined_reg <- lm(G3.x ~ ., data = student_combined)

summary(math_reg)
summary(por_reg)
summary(combined_reg)
```