---
title: "Final Project"
author: "London Wagner"
date: "4/20/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries

```{r}
library(here)
library(tidyverse)
```

## Loading in data

```{r}
student_por <- read_delim(here("data", "student-por.csv"), delim = ";")
```

## Introduction
For our final project we are looking at the student performance data set from the UCI Machine Learning Repository that measures student performance in a Portuguese class based on a variety of predictors. There predictors include but are not limited to:

- student sex
- student age
- previous grades
- access to internet, etc

The person who originally collected this data and performed the study was Paulo Cortez from the University of Minho. 

The purpose of our analysis is determine which variables are significant when predicting a student's final grade in the Portuguese class.




## Running a simple linear regression

```{r}
por_reg <- lm(G3 ~ ., data = student_por)
summary(por_reg)
```

A lot of the predictors do not appear to be significant, so we are going to use some variable selection methods to simplify the model while still maintaing accuracy.

```{r}

```

